Forecasting biodiversity in breeding birds using best practices

### 2018-01-05 manuscript/draft.Rmd

Predictions from the "average" model are thus centered on $\mu$, which could
either be the mean of the raw training richness values, or an output from the
observer model.  This model's confidence intervals have a constant width that
depends on the standard deviation of $\epsilon$, which can either be the
standard deviation of the raw training richness values, or
$\sigma^\mathrm{residual}$ from the observer model; see supplement).

@@

**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
function in the package `forecast`; @forecast1] to represent an array of
different time-series modeling approaches. These models can include an
autoregressive component (as in the "naive" model, but with the possibility of
longer-term dependencies in the underlying process), a moving average component
(where the noise can have serial autocorrelation) and an
integration/differencing component (so that the analysis could be performed on
sequential differences of the raw data, accommodating more complex patterns
including trends). The `auto.arima` function chooses whether to include each of
these components (and how many terms to include for each one) using AICc
[@forecast1]. Since there is no seasonal component to the BBS time-series, we
did not include a season component in these models. Otherwise we used the
default settings for this function (See supplement for details).

@@

In contrast to the single-site models, most attempts to predict species richness
focus on using correlative models based on environmental variables. We tested
three common variants of this approach: direct modeling of species richness;
stacking individual species distribution models; and joint species distribution
models (JSDMs). Following the standard approach, site-level random effects were
not included in these models as predictors, meaning that this approach
implicitly assumes that two sites with identical Bioclim, elevation, and NDVI
values should have identical richness distributions. As above, we included
observer effects and the associated uncertainty by running these models 500
times (once per MCMC sample).

@@

Most models produced confidence intervals that were too narrow, indicating
overconfident predictions (Figure 5C). The random forest-based SDM stack was the
most overconfident model, with only `r numbers$rf_coverage_pct`% of observations
falling inside its 95% confidence intervals. This stacked SDM's narrow
predictive distribution caused it to have notably higher deviance (Figure 5B)
than the next-worst model, even though its point estimates were not unusually
bad in terms of RMSE (5A). As discussed elsewhere [@harris2015], this
overconfidence is a product of the assumption in stacked SDMs that errors in the
species-level predictions are independent. The GBM-based "macroecological" model
and the mistnet JSDM had the best calibrated uncertainty estimates (Figure
5B)and therefore their relative performance was higher in terms of deviance than
in terms of RMSE. The "naive" model was the only model whose confidence
intervals were too wide (Figure 5C), which can be attributed to the rapid rate
at which these intervals expand (Figure 1).

@@

Forecasting is an emerging imperative in ecology; as such, the field needs to
develop and follow best practices for conducting and evaluating ecological
forecasts [@clark2001]. We have used a number of these practices (Box 1) in a
single study that builds and evaluates forecasts of biodiversity in the form of
species richness. The results of this effort are both promising and humbling.
When comparing predictions across sites, many different approaches produce
reasonable forecasts (Figure 3). If a site is predicted to have a high number of
species in the future, relative to other sites, it generally does.  However,
none of the methods evaluated reliably determined how site-level richness
changes over time (Figure 6), which is generally the stated purpose of these
forecasts. As a result, baseline models, which did not attempt to anticipate
changes in richness over time, generally provided the best forecasts for future
biodiversity. While this study is restricted to breeding birds in North America,
its results are consistent with a growing literature on the limits of ecological
forecasting, as discussed below.

The most commonly used methods for forecasting future biodiversity, SDMs and
macroecological models, both produced worse forecasts than time-series models
and simple baselines. This weakness suggests that predictions about future
biodiversity change should be viewed with skepticism unless the underlying
models have been validated temporally, via hindcasting and comparison with
simple baselines. Since site-level richness is relatively stable, spatial
validation is not enough: a model can have high accuracy across spatial
gradients without being able to predict changes over time. This gap between
spatial and temporal accuracy is known to be important for species-level
predictions [@rapacciuolo2012; @oedekoven2017]; our results indicate that it is
substantial for higher-level patterns like richness as well. SDMs' poor temporal
predictions are particularly sobering, as these models have been one of the main
foundations for estimates of the predicted loss of biodiversity to climate
change over the past decade or so [@thomas2004; @thuiller2011; @urban2015]. Our
results also highlight the importance of comparing multiple modeling approaches
when conducting ecological forecasts, and in particular, the value of comparing
results to simple baselines to avoid over-interpreting the information present
in these forecasts [Box 1]. Disciplines that have more mature forecasting
cultures often do this by reporting "forecast skill", i.e., the improvement in
the forecast relative to a simple baseline [@jolliffe2003]. We recommend
following the example of @perretti2013 and adopting this approach in future
ecological forecasting research.

@@

We have only evaluated annual forecasts up to a decade into the future, but
forecasts are often made with a lead time of 50 years or more. These long-term
forecasts are difficult to evaluate given the small number of century-scale
datasets, but are important for understanding changes in biodiversity at some of
the lead times relevant for conservation and management. Two studies have
assessed models of species richness at longer lead times [@algar2009;
@distler2015], but the results were not compared to baseline or time-series
models (in part due to data limitations) making them difficult to compare to our
results directly. Studies on shorter time scales, such as ours, provide one way
to evaluate our forecasting methods without having to wait several decades to
observe the effects of environmental change on biodiversity [@petchey2015;
@tredennick2016; @dietze2016], but cannot fully replace longer-term evaluations
[@tredennick2016]. In general, drivers of species richness can differ at
different temporal scales [@rosenzweig1995; @white2004; @white2007;
@blonder2017], so different methods may perform better for different lead times.
In particular, we might expect environmental and ecological information to
become more important at longer time scales, and thus for the performance of
simple baseline forecasts to degrade faster than forecasts from SDMs and other
similar models. We did observe a small trend in this direction: deviance for
the auto-ARIMA models and for the average baseline grew faster than for two of
the environmental models (the JSDM and the macroecological model), although this
growth was not statistically significant for the average baseline.

@@

Future biodiversity forecasting efforts also need to address the uncertainty
introduced by the error in forecasting the environmental conditions that are
used as predictor variables. In this, and other hindcasting studies, the
environmental conditions for the "future" are known because the data has already
been observed. However, in real forecasts the environmental conditions
themselves have to be predicted, and environmental forecasts will also have
uncertainty and bias. Ultimately, ecological forecasts that use environmental
data will therefore be more uncertain than our current hindcasting efforts, and
it is important to correctly incorporate this uncertainty into our models
[@clark2001; @dietze2017]. Limitations in forecasting future environmental
conditions---particularly at small scales---will present continued challenges
for models incorporating environmental variables, and this may result in a
continued advantage for simple single-site approaches.

@@

The science of forecasting biodiversity remains in its infancy and it is
important to consider weaknesses in current forecasting methods in that context.
In the beginning, weather forecasts were also worse than simple baselines, but
these forecasts have continually improved throughout the history of the field
[@mcgill2012; @silver2012; @bauer2015]. One practice that lead to improvements
in weather forecasts was that large numbers of forecasts were made publicly,
allowing different approaches to be regularly assessed and refined [@mcgill2012;
@silver2012]. To facilitate this kind of improvement, it is important for
ecologists to start regularly making and evaluating real ecological forecasts,
even if they perform poorly, and to make these forecasts openly available for
assessment [@mcgill2012; @dietze2016]. These forecasts should include both
short-term predictions, which can be assessed quickly, and mid- to long-term
forecasts, which can help ecologists to assess long time-scale processes and
determine how far into the future we can successfully forecast [@dietze2016;
@tredennick2016]. We have openly archived forecasts from all six models through
the year 2050 [@zenodoPredict], so that we and others can assess how well they
perform. We plan to evaluate these forecasts and report the results as each new
year of BBS data becomes available, and make iterative improvements to the
forecasting models in response to these assessments.

>>>>

Predictions from the "average" model are thus centered on $\mu$, which could
either be the mean of the raw training richness values, or an output from the
observer model. The confidence iterval width in this model is constant and a 
consequence of the standard deviation $\epsilon$, which can either be the
standard deviation of the raw training richness values, or
$\sigma^\mathrm{residual}$ from the observer model; see supplement).

@@

**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
function in the package `forecast`; @forecast1] to represent an array of
different time-series modeling approaches. These models can include an
autoregressive component (as in the "naive" model, but with the possibility of
longer-term dependencies in the underlying process), a moving average component
(where the noise may be serially autocorrelation) and an
integration/differencing component (so that the analysis could be performed on
sequential differences of the raw data, accommodating more complex patterns
including trends). The `auto.arima` function chooses whether to include each of
these components (and how many terms to include for each one) using AICc
[@forecast1]. Since there is no seasonal component to the BBS time-series, we
did not include a season component in these models. Otherwise we used the
default settings for this function (See supplement for details).

@@

In contrast to the single-site models, most attempts to predict species richness
focus on using correlative models based on environmental variables. We tested
three common variants of this approach: direct modeling of species richness;
stacking individual species distribution models; and joint species distribution
models (JSDMs). Following the standard approach, site-level random effects were
not included in these models as predictors, meaning that this approach
implicitly assumes that two sites with identical Bioclim, elevation, and NDVI
values will have identical richness distributions. As above, we included
observer effects and the associated uncertainty by running these models 500
times (once per MCMC sample).

@@

Most models produced confidence intervals that were too narrow, indicating
overconfident predictions (Figure 5C). The random forest-based SDM stack was the
most overconfident model, with only `r numbers$rf_coverage_pct`% of observations
falling inside its 95% confidence intervals. This stacked SDM's narrow
predictive distribution caused notably greater deviance (Figure 5B)
than the next-worst model, even though its point estimates were not unusually
bad in terms of RMSE (5A). As discussed elsewhere [@harris2015], this
overconfidence is a product of the assumption in stacked SDMs that errors in the
species-level predictions are independent. The GBM-based "macroecological" model
and the mistnet JSDM had the best calibrated uncertainty estimates (Figure
5B)and therefore their relative performance was higher in terms of deviance than
in terms of RMSE. The "naive" model was the only model whose confidence
intervals were too wide (Figure 5C), which can be attributed to the rapid rate
at which these intervals expand (Figure 1).

@@

Forecasting is an emerging imperative in ecology; as such, the field needs to
develop and follow best practices for conducting and evaluating ecological
forecasts [@clark2001]. We used a number of these practices (Box 1) in a
single study that builds and evaluates forecasts of biodiversity in the form of
species richness. The results of this effort are both promising and humbling.
When comparing predictions across sites, many different approaches produce
reasonable forecasts (Figure 3). If a site is predicted to have high species
richness in the future, relative to other sites, it generally does.  However,
none of the methods evaluated reliably determined how site-level richness
changes over time (Figure 6), which is generally the stated purpose of these
forecasts. As a result, baseline models, which did not attempt to anticipate
changes in richness over time, generally provided the best forecasts for future
biodiversity. While this study is restricted to breeding birds in North America,
its results are consistent with a growing literature on the limits of ecological
forecasting, as discussed below.

The most commonly used methods for forecasting future biodiversity, SDMs and
macroecological models, both produced worse forecasts than time-series models
and simple baselines. This weakness suggests that predictions about future
biodiversity change should be viewed with skepticism unless the underlying
models are temporally validated, via hindcasting and comparison with
simple baselines. Since site-level richness is relatively stable, spatial
validation is not enough: a model be accurate across spatial gradients while
being unable to predict changes over time. This gap between
spatial and temporal accuracy is important for species-level
predictions [@rapacciuolo2012; @oedekoven2017]; our results indicate that it is
substantial for higher-level patterns like richness as well. The poor quality 
temporal predictions of SDM's are particularly sobering because of their importance
over the last decade or so for predicting biodiversity loss as a response to 
climate change [@thomas2004; @thuiller2011; @urban2015]. Our
results also highlight the importance of comparing multiple modeling approaches
when conducting ecological forecasts, and in particular, the value of comparing
results to simple baselines to avoid over-interpreting the information present
in these forecasts [Box 1]. Disciplines with more mature forecasting
cultures often do this by reporting "forecast skill", i.e., the improvement in
the forecast relative to a simple baseline [@jolliffe2003]. We recommend
following the example of @perretti2013 and adopting this approach in future
ecological forecasting research.

@@

We only evaluated annual forecasts up to a decade into the future, but
forecasts are often made with a lead time of 50 years or more. These long-term
forecasts are difficult to evaluate given the small number of century-scale
datasets, but are important for understanding changes in biodiversity at some of
the lead times relevant for conservation and management. Two studies
assessed models of species richness at longer lead times [@algar2009;
@distler2015], but the results were not compared to baseline or time-series
models (in part due to data limitations) making them difficult to compare to our
results directly. Studies on shorter time scales, such as ours, provide one way
to evaluate our forecasting methods without having to wait several decades to
observe the effects of environmental change on biodiversity [@petchey2015;
@tredennick2016; @dietze2016], but cannot fully replace longer-term evaluations
[@tredennick2016]. In general, drivers of species richness can differ at
different temporal scales [@rosenzweig1995; @white2004; @white2007;
@blonder2017], so different methods may perform better for different lead times.
In particular, we might expect environmental and ecological information to
become more important at longer time scales, and thus for the performance of
simple baseline forecasts to degrade faster than forecasts from SDMs and other
similar models. We did observe a small trend in this direction: deviance for
the auto-ARIMA models and for the average baseline grew faster than for two of
the environmental models (the JSDM and the macroecological model), although this
growth was not statistically significant for the average baseline.

@@

Future biodiversity forecasting efforts also need to address the uncertainty
introduced by the error in forecasting the environmental conditions that are
used as predictor variables. In this, and other hindcasting studies, the
environmental conditions for the "future" are known because the data has already
been observed. However, in real forecasts the environmental conditions
themselves must be predicted, and environmental forecasts will also have
uncertainty and bias. Ultimately, ecological forecasts that use environmental
data will therefore be more uncertain than our current hindcasting efforts, and
it is important to correctly incorporate this uncertainty into our models
[@clark2001; @dietze2017]. Limitations in forecasting future environmental
conditions---particularly at small scales---will present continued challenges
for models incorporating environmental variables, and this may result in a
continued advantage for simple single-site approaches.

@@

The science of forecasting biodiversity remains in its infancy and it is
important to consider weaknesses in current forecasting methods in that context.
In the beginning, weather forecasts were also worse than simple baselines, but
these forecasts have continually improved throughout the history of the field
[@mcgill2012; @silver2012; @bauer2015]. One practice that lead to improvements
in weather forecasts was that large numbers of forecasts were made publicly,
allowing different approaches to be regularly assessed and refined [@mcgill2012;
@silver2012]. To facilitate this kind of improvement, it is important for
ecologists to start regularly making and evaluating real ecological forecasts,
even if they perform poorly, and to make these forecasts openly available for
assessment [@mcgill2012; @dietze2016]. These forecasts should include both
short-term predictions, which can be assessed quickly, and mid- to long-term
forecasts, which can help ecologists to assess long time-scale processes and
determine how far into the future we can successfully forecast [@dietze2016;
@tredennick2016]. We openly archived forecasts from all six models through
the year 2050 [@zenodoPredict], so that we and others can assess how well they
perform. We plan to evaluate these forecasts and report the results as each new
year of BBS data becomes available, and make iterative improvements to the
forecasting models in response to these assessments.
